Detecting and quantifying parasite-induced host mortality from intensity data: method comparisons and limitations

### 2015-08-04 docs/response_to_reviewers.md

## Response to Reviewers

### Reviewer 1

We thank Reviewer 1 for the helpful suggestions. We agree with Reviewer 1 that host-parasite distributions are dynamic and that it would be interesting to apply these models to

### Reviewer 2

We thank Reviewer 2 for their helpful feedback and address each of their points in turn.

1. The major critique provided by Reviewer 2 was that we were too negative about our results in the Discussion and the Abstract. We agree with this comment and have reworded the Abstract and the Discussion to reflect why we think .  That being said, we do still feel it is important to highlight the limitations of

### 2015-08-04 docs/parasite_mortality_manuscript.tex
Beginning to respond to reviewers

Parasites can significantly impact animal populations by changing host
behavior, reproduction and survival. Detecting and quantifying these impacts is
critical for understanding disease dynamics and managing wild animal
populations. However, for wild hosts infected with macroparasites, it is
notoriously difficult to quantify the fatal parasite load and number of animals
that have died due to disease. When ethical or logistical constraints prohibit
experimental determination of these values, examination of parasite intensity
and distribution data may offer an alternative solution. In this study we
introduce a novel method for using intensity data to detect and quantify
parasite-induced mortality in wildlife populations. We use simulations to show
that this method is more reliable than previously proposed methods while
providing quantitative estimates of parasite-induced mortality from empirical
data that are consistent with previously published qualitative estimates.
However, we stress that this method, and all techniques that estimate parasite-induced mortality from intensity data alone, have several critical assumptions
that limit their applicability in the real world. Due to these limitations,
these methods should only be used as an exploratory tool to inform more
rigorous studies of parasite-induced host mortality.

@@

Generally, a larger $b$ leads to a more rapid decline in the probability of host survival as parasite intensity increases, with the maximum rate of decline having a value of $b / 4$ (\emph{SI} 2). $b$ is in many ways analogous to the pathogenicity parameter ($\alpha$) in classic macroparasite models that gives the parasite intensity dependent host death rate. \citep{AndersonandMay1978,Isham1995}. When $b$ is held constant, a larger $a$ allows for hosts to tolerate larger parasite intensities before experiencing parasite-induced mortality. More specifically, for every one unit increase in $a$ the log parasite intensity at which any percent of hosts survive (e.g. 99\% of hosts survive) increases by $1 / b$ (\emph{SI} 2).

@@

We thank the Briggs Lab group and Theoretical Ecology group at UCSB for helpful feedback.  MW was supported by National Science Foundation Graduate Research Fellowship (Grant No. DGE 1144085) and the UC Regents CB was supported by NIH grant 1R01GM109499 from the Ecology of Infectious Disease program.

>>>>

Parasites can significantly impact animal populations by changing host
behavior, reproduction and survival. Detecting and quantifying these impacts is
critical for understanding disease dynamics and managing wild animal
populations. However, for wild hosts infected with macroparasites, it is
notoriously difficult to quantify the fatal parasite load and number of animals
that have died due to disease. When ethical or logistical constraints prohibit
experimental determination of these values, examination of parasite intensity
and distribution data may offer an alternative solution. In this study we
introduce a novel method for using intensity data to detect and quantify
parasite-induced mortality in wildlife populations. We use simulations to show
that this method is more reliable than previously proposed methods while
providing quantitative estimates of parasite-induced mortality from empirical
data that are consistent with previously published qualitative estimates.
However, this method, and all techniques that estimate parasite-induced
mortality from intensity data alone, have several important assumptions that
must be scrutinized before applying them to real-world data. Given that these
assumptions are met, our method is new exploratory tool that can help inform
more rigorous studies of parasite-induced host mortality.

@@

Generally, a larger $b$ leads to a more rapid decline in the probability of host survival as parasite intensity increases, with the maximum rate of decline having a value of $b / 4$ (\emph{SI} 2). $b$ is in many ways analogous to the pathogenicity parameter ($\alpha$) in classic macroparasite models that gives the parasite intensity dependent host death rate \citep{AndersonandMay1978,Isham1995}. When $b$ is held constant, a larger $a$ allows for hosts to tolerate larger parasite intensities before experiencing parasite-induced mortality. More specifically, for every one unit increase in $a$ the log parasite intensity at which any percent of hosts survive (e.g. 99\% of hosts survive) increases by $1 / b$ (\emph{SI} 2).

@@

We thank the Briggs Lab group, Kuris/Lafferty Lab group, Theoretical Ecology group at UCSB, and two anonymous reviewers for helpful feedback.  MW was supported by National Science Foundation Graduate Research Fellowship (Grant No. DGE 1144085) and the UC Regents. CB was supported by NIH grant 1R01GM109499 from the Ecology of Infectious Disease program.

### 2015-08-04 docs/supplementary_material.tex
Beginning to respond to reviewers

The second step of the Adjei Method is to make the assumption that infection, host mortality, and sampling occur in that order and are temporally \cite{Adjei1986}. Next, \citeauthor{Adjei1986} assume that the host survival function follows the logistic form

@@

\begin{align*}
    \log{\frac{p}{1 - p}} &= a_1 - b \log{x_1}\\
    - \log{\frac{p}{1 - p}} &= a_1 - b \log{x_2}\\
    0 &= a_1 - a_2 - b\log{x_1} + b\log{x_2}\\
    a_2 - a_1 &= b(\log{x_2} - \log{x_1})
\end{align*}
If $a_2 - a_1 = 1$, than the change in log parasite intensity at which $p$ percentage of hosts survive is $1 / b$.

>>>>

The second step of the Adjei Method is to make the assumption that infection, host mortality, and sampling occur in that order and are temporally separate \citep{Adjei1986}. Next, \citeauthor{Adjei1986} assume that the host survival function follows the logistic form

@@

\begin{align*}
    \log{\frac{p}{1 - p}} &= a_1 - b \log{x_1}\\
    - \log{\frac{p}{1 - p}} &= a_2 - b \log{x_2}\\
    0 &= a_1 - a_2 - b\log{x_1} + b\log{x_2}\\
    a_2 - a_1 &= b(\log{x_2} - \log{x_1})
\end{align*}
If $a_2 - a_1 = 1$, than the change in log parasite intensity at which $p$ percentage of hosts survive is $1 / b$.
